---
title: "EDA and Data Cleaning on the `mtcars` Dataset"
author: "[Masud Rahman](https://masud90.github.io/)"
format:
  gfm:
     html-math-method: 
       method: webtex
       url: https://latex.codecogs.com/png.image?%5Cbg_black&space;
---

```{r setup, include=FALSE}
# Set global options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load necessary libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(GGally)
library(reshape2)
```

[![GitHub](https://img.shields.io/badge/GitHub-100000?style=flat&logo=github&logoColor=white)](https://github.com/masud90/)
[![LinkedIn](https://img.shields.io/badge/LinkedIn-0077B5?style=flat&logo=linkedin&logoColor=white)](https://linkedin.com/in/Masud90)
[![Twitter Follow](https://img.shields.io/twitter/follow/masudtweets?style=social&logo=twitter)](https://twitter.com/masudtweets)

# Introduction

The `mtcars` dataset is a well-known dataset in R that contains data extracted from the 1974 Motor Trend US magazine. The dataset comprises various automobile design and performance aspects for 32 cars, including miles per gallon, number of cylinders, horsepower, weight, and more. In this report, we will perform data cleaning and exploratory data analysis (EDA) to uncover the underlying patterns in the data.

## Load the mtcars dataset and Inspect the first few rows of the dataset

```{r}
data("mtcars")
head(mtcars)
```

## Summary statistics

```{r}
summary(mtcars)
```

# Pre-processing

## Check for missing values in the dataset

```{r}
missing_values <- sum(is.na(mtcars))
if (missing_values == 0) {
  print("There are no missing values in the dataset.")
} else {
  print(paste("There are", missing_values, "missing values in the dataset."))
}
```

## Inspect the data types of all columns

```{r}
str(mtcars)
```

## Convert numerical features to factors for categorical analysis

```{r}
mtcars$cyl <- as.factor(mtcars$cyl)
mtcars$vs <- as.factor(mtcars$vs)
mtcars$am <- as.factor(mtcars$am)
mtcars$gear <- as.factor(mtcars$gear)
mtcars$carb <- as.factor(mtcars$carb)
```

## Check the structure again to confirm data type changes

```{r}
str(mtcars)
```

# Exploratory Data Analysis

## Histogram: Miles per Gallon

```{r}
#| fig-cap: "Histogram showing the distribution of miles per gallon (mpg) across the cars."
ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(binwidth = 2, fill = "skyblue", color = "black") +
  labs(x = "Miles Per Gallon", y = "Frequency")
```

## Boxplot of MPG by Number of Cylinders

```{r}
#| fig-cap: "Boxplot depicting the variation in miles per gallon (mpg) and cylinders."
ggplot(mtcars, aes(x = cyl, y = mpg, fill = cyl)) +
  geom_boxplot() +
  scale_fill_brewer(palette = "Set3") +
  labs(x = "Number of Cylinders", y = "Miles Per Gallon")
```

## Scatter Plot of Horsepower vs. Miles Per Gallon

```{r}
#| fig-cap: "Scatter plot showing the relationship between horsepower and miles per gallon (mpg)."
ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point() +
  geom_smooth(method = "lm", color = "darkblue", se = FALSE) +
  labs(x = "Horsepower", y = "Miles Per Gallon")
```

## Pairwise Plot of All Variables

```{r}
#| fig-cap: "Pairwise plot illustrating the relationships between all numeric variables."
ggpairs(mtcars, columns = 1:7, ggplot2::aes(color = cyl))
```

## Correlation Heatmap

```{r}
#| fig-cap: "Heatmap showing the Pearson correlation coefficients."
# Select numeric columns
mtcars_numeric <- mtcars %>%
  select_if(is.numeric)

# Calculate correlation matrix
cor_matrix <- round(cor(mtcars_numeric), 2)

# Reshape the correlation matrix for plotting
cor_melted <- melt(cor_matrix)

# Plot the correlation heatmap
ggplot(cor_melted, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "red", high = "blue", mid = "white", 
                       midpoint = 0, limit = c(-1, 1), space = "Lab", 
                       name = "Pearson\nCorrelation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, 
                                   size = 12, hjust = 1)) +
  coord_fixed()
```
